---
title: "Diamonds"  
author: "Prera & Agrawal"     

# Output formats and their specific options
format:
  pdf:
    toc: true
    number-sections: true
    pdf-engine: lualatex
    keep-tex: true
  html:
    toc: true
    theme: cosmo
    code-fold: true
  docx:
    reference-doc: custom.docx
    
# Global execution behavior
code-line-numbers: false
colorlinks: true
urlcolor: purple
citecolor: purple
execute:      
  message: false                                
  warning: false
  include: false                         
  freeze: auto   

# Appearance and structure
fontsize: 11pt                                  
highlight-style: tango                          
toc-title: Contents    
toccolor: blue
toc-depth: 3                                    
number-depth: 3    

# Figures
fig-align: center
fig-cap-location: bottom

# Citations
cite-method: citeproc                           
# bibliography: references.bib                    
# csl: apa.csl                                    

# Page geometry (PDF)
geometry:       
  - top=0.75in                                  
  - left=0.75in                                 
  - bottom=0.75in                               
  - right=0.75in                                

# Additional LaTeX control
header-includes:                                
  - \usepackage{hyperref}
  -  \newenvironment{customabstract}{
        \begin{center}
        {\bfseries Abstract}\par\vspace{0.5em}
        \begin{minipage}{0.8\linewidth}
      }{
        \end{minipage}
        \end{center}
      }
---

\newpage

\begin{customabstract}
This is my abstract
\end{customabstract}

\newpage

# Introduction

# Exploratory Data Analysis

```{r}
### important packages
library(tidyverse)
library(patchwork)
library(dplyr)
```

```{r}
diamonds <- read_csv("diamonds.csv")
```

```{r}
diamonds |> glimpse()
```

```{r}
#| include: true
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: "center"
#| fig-pos: "H"
#| fig-cap: "Price vs Diamonds"
#| label: fig-price-vs-diamond

diamonds_sampled <- diamonds |>
  slice_sample(n = 20000)

diamonds_sampled <- diamonds_sampled |>
  mutate(volume = x * y * z)

g1 <- diamonds |> 
  ggplot(aes(x = carat, y = price )) +
  geom_point(size = 0.0001)+
  coord_cartesian(xlim = c(0, 3.5)) +
  theme_bw()

g2 <- diamonds |>
  ggplot(aes(x =carat, y = sqrt(price))) +
  geom_point(size = 0.0001) +
  coord_cartesian(xlim = c(0, 3.5)) +
  theme_bw()

g3 <- diamonds |>
  ggplot(aes(x =carat, y = log(price))) +
  geom_point(size = 0.0001) +
  coord_cartesian(xlim = c(0, 3.5)) +
  theme_bw()
    
(g1 / g3) | g2
```


Firstly, we tried to understand the relationship of Diamond prices with the predictors we have in the dataset. The first obvious predictor to consider was carats, as popularized by advertisements on the Fifth Avenue by De Beers, "Bigger is Better." (Cite the book Many Meanings of Diamonds, Susan Falls) We guessed initially that it would not be a linear relationship, thus, we immediately applied a logarithmic transformation. To our dismay, the graph just looked reflected and after an hour of other forms of data exploration, we tried applying a square root transformation, and the graph became immediately linear. 


```{r}
#| include: true
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: "center"
#| fig-pos: "H"
#| fig-cap: "Price vs volume"
#| label: fig-price-vs-volume

g1 <- diamonds_sampled |>
  ggplot(aes(x = volume, y = sqrt(price))) +
  geom_point(size = 0.0001) +
  coord_cartesian(xlim = c(0, 1000)) +
  theme_bw()

g2 <- diamonds_sampled |>
  ggplot(aes(x = x^3, y = sqrt(price))) +
  geom_point(size = 0.0001) +
  coord_cartesian(xlim = c(0, 1000)) +
  theme_bw()

g3 <- diamonds_sampled |>
  ggplot(aes(x = y^3, y = sqrt(price))) +
  geom_point(size = 0.0001) +
  coord_cartesian(xlim = c(0, 1000)) +
  theme_bw()

g4 <- diamonds_sampled |>
  ggplot(aes(x = z^3, y = sqrt(price))) +
  geom_point(size = 0.0001) +
  coord_cartesian(xlim = c(0, 1000)) +
  theme_bw()

(g1 + g2) / (g3 + g4)
```

Maybe talk about how the slopes compare to each other. The first volume graph is obvious, but the difference between the slopes of the other 3 seems to be quite different. Which means with growing depth, the diamond price increases faster as compared to the increase in length or width.

Apart from that, it is pretty fascinating that the price is just correlated exponentially with the volume, and thus also exponentially with the rest of the items.

If we were to strictly stick to a linear regression, we should not be caring about x, y, and z, we may lose some information but we need to use volume because we cannot cube the x, y, and z.


```{r}
#| include: true
#| echo: false
#| fig-width: 9
#| fig-height: 5
#| fig-align: "center"
#| fig-pos: "H"
#| fig-cap: "Price vs Diamonds"
#| label: fig-price-vs-diamond-vs-cut

g1 <- diamonds_sampled |> 
  ggplot(aes(x = carat, y = price, color = cut)) +
  geom_point(size = 0.0001)+
  theme_bw()

g2 <- diamonds_sampled |>
  ggplot(aes(x =carat, y = log(price), color = cut)) +
  geom_point(size = 0.0001) +
  theme_bw()

g3 <- diamonds_sampled |>
  ggplot(aes(x = carat, y = sqrt(price), color = cut)) +
  geom_point(size = 0.001) +
  geom_smooth(method = "lm", se = FALSE) +
  coord_cartesian(ylim = c(20, 130), xlim = c(0, 3.5)) +
  theme_bw() +
  labs(title = "Linear Regression on Price, Carats and Cut of Diamonds.",
  x = "Price",
  y = "Carat")
    
g3
```

```{r}
#| include: true
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| fig-align: "center"
#| fig-pos: "H"
#| fig-cap: "Histogram of Diamond Prices."
#| label: fig-prices-hist

diamonds |>
  ggplot(aes(x = price)) +
  geom_histogram(fill = "lightblue", color = "black", bins = 150) +
  theme_bw() +
  labs(
    title = "Histogram of Diamond Prices",
    x = "Price",
    y = "Count"
  )


```

```{r}
#| include: true
#| echo: false
#| fig-width: 9
#| fig-height: 4
#| fig-align: "center"
#| fig-pos: "H"
#| fig-cap: "Histogram of Diamond depth and price"
#| label: fig-prices-depth-scatt

diamonds |>
    ggplot(aes(x = depth, y = price )) +
    geom_point(size = 0.0001, color= "pink") +
    theme_bw()
    labs(
    title = "Histogram of Diamond Prices vs depth",
    x = "depth",
    y = "price"
     )
```

```{r}
### data example
data(mtcars)
mtcars |> glimpse()
```

# Methods

@fig-price-vs-diamond

# Conclusion

# References
